---
title: "Overview and Setup"
output:
  html_document:
    code_download: false
---


## Prerequisites

- A computer
- R at least version 2.1
- R packages
    - sandwich, geepack

## Installation

To get the most out of this tutorial, we recommend installing the development version of `eventglm`. Choose one of the following options: 


### To install from the binaries on `r-universe` [Option 1]

```{r, eval = FALSE, class.source="bg-success"}
install.packages("eventglm", repos = c(
  ropensci = 'https://sachsmc.r-universe.dev',
  CRAN = 'https://cloud.r-project.org'))
```


### **If you are setup to compile R packages from source** [Option 2] 

```{r, eval = FALSE, class.source = "bg-warning"}
remotes::install_github("sachsmc/eventglm")
```


### Download and install the binary for your system [Option 3]

- Windows: https://sachsmc.r-universe.dev/bin/windows/contrib/4.2/eventglm_1.4.0.zip
- OSX: https://sachsmc.r-universe.dev/bin/macosx/contrib/4.2/eventglm_1.4.0.tgz


### The stable version of the package can be installed from CRAN

```{r, eval = FALSE, class.source= "bg-success"}
install.packages("eventglm")
```


## The `eventglm` package

The github page for `eventglm` is https://github.com/sachsmc/eventglm . Documentation and vignettes are available in R and at https://sachsmc.github.io/eventglm . 

To cite eventglm in publications use:

> Sachs MC, Gabriel EE (2022). “Event History Regression with Pseudo-Observations: Computational
> Approaches and an Implementation in R.” _Journal of Statistical Software_, *102*(9), 1-34. doi:
> 10.18637/jss.v102.i09 (URL: https://doi.org/10.18637/jss.v102.i09).

A BibTeX entry for LaTeX users is

```
  @Article{,
    title = {Event History Regression with Pseudo-Observations: Computational Approaches and an Implementation in {R}},
    author = {Michael C. Sachs and Erin E. Gabriel},
    journal = {Journal of Statistical Software},
    year = {2022},
    volume = {102},
    number = {9},
    pages = {1--34},
    doi = {10.18637/jss.v102.i09},
  }
```


