---
title: "Joint modelling multiple endpoints"
output:
  html_document:
    code_folding: show
bibliography: refs.bib
---


### Learning objectives {.alert .alert-success}

In this lesson you will 

1. Write a function to perform simultaneous inference for covariate effects on multiple outcomes in the competing risks setting. 

## Setup

Library the `survival` package and then the `eventglm` package, and then the `geepack` package. 


```{r}
library(survival)
library(eventglm)
library(geepack)
```

## Joint models

Consider again the `mgus2` dataset. This is a competing risks setting, with the two possible causes of failure death and pcm. 

```{r}
crfit <- survfit(Surv(etime, event) ~ sex, eventglm::mgus2)
summary(crfit, times = 120)
print(crfit, rmean = 120)
plot(crfit, col=1:2,  noplot="",
     lty=c(3,3,2,2,1,1), lwd=2, xscale=12,
     xlab="Years post diagnosis", ylab="P(state)")
legend(240, .65, c("Female, death", "Male, death", "malignancy", "(s0)"),
       lty=c(1,1,2,3), col=c(1,2,1,1), bty='n', lwd=2)
abline(v = 120, lty = 2)
```

This is the _competing risks_ setting, which can be described with the multi-state model in the figure below. 

```{r}
connect <- matrix(0, nrow = 3, ncol = 3, 
                  dimnames = lapply(1:2, \(i) c("MGUS", "PCM", "Death")))
connect[1, 2:3] <- 1
statefig(c(1, 2), connect)
```


Suppose we are interested in the effect of the mspike variable on __either pcm or death__. We fit the following models
```{r}
mgus2 <- subset(mgus2, !is.na(mspike))

mgcipcm <- cumincglm(Surv(etime, event) ~ mspike, cause = "pcm", time = 120, 
                   data = mgus2)
mgcideath <- cumincglm(Surv(etime, event) ~ mspike, cause = "death", time = 120, 
                   data = mgus2)
```
which gives us two coefficients for `mspike`. 


### Exercise {.alert .alert-warning}

Write a function to perform simultaneous inference on the two coefficients. That is, find a way to test the null hypothesis that both mspike coefficients (in the pcm model and the death model) are equal to 0. 


<details>
<summary><strong>Solution 1</strong></summary>

This solution involves stacking the estimating equations for both models, and the refitting with an indicator for the outcome type. 

```{r}
jointfit <- function(fit1, fit2) {
  
  bform <- update.formula(fit1$formula, .PO ~ .)
  
  dset1 <- get_all_vars(bform[-2], fit1$data)
  dset1$.PO <- fit1$y
  dset1$model <- "fit1"
  dset1$id <- 1:nrow(dset1)
  
  dset2 <- get_all_vars(bform[-2], fit2$data)
  dset2$.PO <- fit2$y
  dset2$model <- "fit2"
  dset2$id <- 1:nrow(dset2)
  
  dset <- rbind(dset1, dset2)
  
  nform <- update.formula(bform, ~ .:model + model - 1)
  
  geefit <- geese(nform, data = dset[order(dset$id),], id = id)
  
  geefit
  
}

jointm <- jointfit(mgcipcm, mgcideath)


mstat <- c(t(jointm$beta[3:4]) %*% solve(jointm$vbeta[3:4, 3:4]) %*% jointm$beta[3:4])
pchisq(mstat, df = 2, lower.tail = FALSE)
```
</details>


<details>
<summary><strong>Solution 2</strong></summary>

This solution uses the bootstrap to estimate the covariance matrix of the two coefficients. 

```{r, cache=TRUE}
B <- 2000
beta.boot <- matrix(NA, nrow = B, ncol = 2)

for(j in 1:B){
 
  mgus2.j <- mgus2[sample(1:nrow(mgus2), nrow(mgus2), replace = TRUE),]
  
  mgcipcm.j <- cumincglm(Surv(etime, event) ~ mspike, cause = "pcm", time = 120, 
                   data = mgus2.j)
  mgcideath.j <- cumincglm(Surv(etime, event) ~ mspike, cause = "death", time = 120, 
                   data = mgus2.j)
  
  beta.boot[j,] <- c(mgcipcm.j$coefficients[2], mgcideath.j$coefficients[2])

}

mstat <- c(mgcipcm$coefficients[2], mgcideath$coefficients[2]) %*% 
  solve(cov(beta.boot)) %*% c(mgcipcm$coefficients[2], mgcideath$coefficients[2])
pchisq(mstat, df = 2, lower.tail = FALSE)
```
</details>


